---
title: "Token Usage & Cost Estimate"
date: today
format: gfm
execute:
  echo: false
  warning: false
params:
  token_csv: "token_counts.csv"
  encoding: "cl100k_base"
  project: ""
  cost_in_per_million: 1.5
  cost_out_per_million: 9.0
  assumed_output_tokens: 7000
  batch_discount: 0.5
---

```{r}
#| label: setup
library(readr)
library(dplyr)
library(ggplot2)
library(scales)

counts <- read_csv(params$token_csv, show_col_types = FALSE)

n_articles <- nrow(counts)
total_in   <- sum(counts$n_tokens_in)
out_each   <- params$assumed_output_tokens
cost_in    <- params$cost_in_per_million
cost_out   <- params$cost_out_per_million

# Fixed input cost; output cost scales with the (uncertain) output assumption.
bd            <- params$batch_discount
cost_in_total <- total_in / 1e6 * cost_in
scenario <- tibble(
  Scenario         = c("Low (0.5x)", "Expected", "High (1.5x)"),
  `Output/article` = round(out_each * c(0.5, 1, 1.5)),
  `Output cost`    = (n_articles * out_each * c(0.5, 1, 1.5)) / 1e6 * cost_out,
  `Input cost`     = cost_in_total,
  `Total`          = cost_in_total + (n_articles * out_each * c(0.5, 1, 1.5)) / 1e6 * cost_out
) |>
  mutate(`Batch total` = `Total` * bd)
```

`r if (nzchar(params$project)) paste0("**Project:** ", params$project, "  ")`
**Articles:** `r comma(n_articles)` &nbsp;|&nbsp; **Encoding:** `r params$encoding` &nbsp;|&nbsp;
**Prices:** $`r sprintf("%.2f", cost_in)`/M in, $`r sprintf("%.2f", cost_out)`/M out &nbsp;|&nbsp;
**Assumed output:** `r comma(out_each)` tokens/article

## Token distribution

```{r}
#| label: fig-density
#| fig-cap: "Density of INPUT tokens per article. Dashed lines mark common context-window sizes."
#| fig-width: 9
#| fig-height: 5
ctx <- c(4096, 8192, 16384, 32768, 65536)
ggplot(counts, aes(x = n_tokens_in)) +
  geom_density(fill = "#4A90D9", alpha = 0.6, colour = "#2C5F8A") +
  geom_vline(xintercept = ctx, linetype = "dashed", colour = "red", alpha = 0.5) +
  annotate("text", x = ctx, y = Inf, label = paste0(ctx / 1024, "K"),
           vjust = 1.5, hjust = -0.15, size = 3, colour = "red") +
  scale_x_continuous(labels = comma) +
  labs(x = "Input tokens per article", y = "Density") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: tbl-summary
q <- quantile(counts$n_tokens_in, c(0.5, 0.9, 0.95, 0.99))
knitr::kable(
  tibble(
    Metric = c("Total input tokens", "Mean", "Median", "90th pct", "95th pct", "99th pct", "Max"),
    Value  = comma(round(c(total_in, mean(counts$n_tokens_in), q[1], q[2], q[3], q[4],
                           max(counts$n_tokens_in))))
  ),
  caption = "Input-token summary."
)
```

## Estimated cost

Input tokens are measured, so the **input cost is fixed**. Output tokens are an
assumption, so the range below flexes the assumed output per article from 0.5x
to 1.5x of `r comma(out_each)`.

```{r}
#| label: tbl-cost
knitr::kable(
  scenario |>
    mutate(across(c(`Output cost`, `Input cost`, `Total`, `Batch total`), ~ dollar(.x, accuracy = 0.01))),
  align = "lrrrrr",
  caption = sprintf("Estimated cost by output-token assumption (batch = %.0f%% off).", (1 - bd) * 100)
)
```

::: {.callout-note}
Expected total: **`r dollar(scenario$Total[2], accuracy = 0.01)` standard**
/ **`r dollar(scenario$"Batch total"[2], accuracy = 0.01)` batch**
(standard range `r dollar(scenario$Total[1], accuracy = 0.01)` –
`r dollar(scenario$Total[3], accuracy = 0.01)`).
Per-article expected cost: median
**`r dollar(median(counts$cost_total_usd), accuracy = 0.0001)`**,
max `r dollar(max(counts$cost_total_usd), accuracy = 0.0001)`.
:::

```{r}
#| label: fig-cost-density
#| fig-cap: "Density of estimated per-article cost (at the expected output assumption)."
#| fig-width: 9
#| fig-height: 4
ggplot(counts, aes(x = cost_total_usd)) +
  geom_density(fill = "#5CB85C", alpha = 0.6, colour = "#3D8B3D") +
  scale_x_continuous(labels = dollar_format(accuracy = 0.001)) +
  labs(x = "Estimated cost per article (USD)", y = "Density") +
  theme_minimal(base_size = 12)
```
