---
title: "Dimensional–Categorical Stance: Descriptive Analysis"
author: "articLLMate"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(here)
library(scales)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title    = element_text(face = "bold"),
      strip_text    = element_text(face = "bold"),
      legend.position = "bottom"
    )
)
```

# 1. Load Data

Point this at your 500-article run output folder.

```{r load-data}
# ---- CHANGE THIS to your run folder name ----
project_name <- "YOUR_PROJECT_NAME"

output_dir   <- here("output", project_name)
results_rds  <- file.path(output_dir, paste0(project_name, "_full_results.rds"))

# Load the full results object (has META + RESULTS per article)
full <- readRDS(results_rds)
cat(sprintf("Loaded %d articles from %s\n", length(full), results_rds))

# Also load the CSVs for convenience
main_df   <- read_csv(file.path(output_dir, paste0(project_name, "_main.csv")),   show_col_types = FALSE)
detail_df <- read_csv(file.path(output_dir, paste0(project_name, "_detail.csv")), show_col_types = FALSE)
```

## Extract publication year from META

```{r extract-year}
year_lookup <- tibble(
  filename = names(full),
  year = map_int(full, function(art) {
    issued <- art$META$issued
    if (!is.null(issued) && !is.na(issued) && nchar(issued) >= 4) {
      return(as.integer(substr(issued, 1, 4)))
    }
    # Fallback: try published.print or published.online
    for (field in c("published.print", "published.online")) {
      val <- art$META[[field]]
      if (!is.null(val) && !is.na(val) && nchar(val) >= 4) {
        return(as.integer(substr(val, 1, 4)))
      }
    }
    return(NA_integer_)
  })
)

# Join year to main data
main_df <- main_df |> left_join(year_lookup, by = "filename")
cat(sprintf("Year available for %d / %d articles\n", sum(!is.na(main_df$year)), nrow(main_df)))
```

# 2. Sample Overview

```{r sample-overview}
status_counts <- main_df |> count(status, sort = TRUE)
knitr::kable(status_counts, caption = "Article processing status")
```

```{r filter-processed}
# Keep only processed + relevant articles for analysis
df <- main_df |>
  filter(status == "PROCESSED",
         relevance_check_is_relevant == TRUE)

cat(sprintf("Analysing %d relevant, processed articles (of %d total)\n", nrow(df), nrow(main_df)))
```

## Articles by year

```{r articles-by-year}
df |>
  filter(!is.na(year)) |>
  count(year) |>
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "#2c7fb8") +
  labs(title = "Number of articles by publication year",
       x = "Year", y = "Count")
```

## Articles by study design

```{r by-study-design}
df |>
  count(study_design, sort = TRUE) |>
  filter(study_design != "N/A") |>
  ggplot(aes(x = reorder(study_design, n), y = n)) +
  geom_col(fill = "#7fcdbb") +
  coord_flip() +
  labs(title = "Articles by study design", x = NULL, y = "Count")
```

## Articles by sample type

```{r by-sample-type}
df |>
  count(sample_type, sort = TRUE) |>
  filter(sample_type != "N/A") |>
  ggplot(aes(x = reorder(sample_type, n), y = n)) +
  geom_col(fill = "#41b6c4") +
  coord_flip() +
  labs(title = "Articles by sample type", x = NULL, y = "Count")
```

## Constructs examined

```{r constructs}
constructs <- detail_df |>
  filter(detail_table == "relevance_check.psychopathologies") |>
  select(filename, construct_label, construct_label_normalised, construct_type,
         personality_disorder, operational_direction, theoretical_direction)

cat(sprintf("%d constructs across %d articles\n",
            nrow(constructs), n_distinct(constructs$filename)))

constructs |>
  count(construct_label_normalised, sort = TRUE) |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(construct_label_normalised, n), y = n)) +
  geom_col(fill = "#253494") +
  coord_flip() +
  labs(title = "Top 20 psychopathology constructs",
       x = NULL, y = "Count")
```

```{r construct-type}
constructs |>
  count(construct_type, sort = TRUE) |>
  ggplot(aes(x = reorder(construct_type, n), y = n)) +
  geom_col(fill = "#41ae76") +
  coord_flip() +
  labs(title = "Construct types", x = NULL, y = "Count")
```


# 3. Overall Classification

```{r overall-classification}
df |>
  count(overall_classification_classification, sort = TRUE) |>
  ggplot(aes(x = reorder(overall_classification_classification, n), y = n)) +
  geom_col(fill = "#d95f0e") +
  coord_flip() +
  labs(title = "Overall classification",
       x = NULL, y = "Count")
```

```{r classification-confidence}
df |>
  ggplot(aes(x = overall_classification_classification,
             y = overall_classification_confidence)) +
  geom_boxplot(fill = "#fec44f", alpha = 0.7) +
  labs(title = "Classification confidence by category",
       x = NULL, y = "Confidence (0–1)")
```


# 4. Dimensional vs Categorical Ratings

## Theoretical stance

```{r theoretical-scatter}
df |>
  ggplot(aes(x = theoretical_stance_categorical_rating,
             y = theoretical_stance_dimensional_rating,
             colour = overall_classification_classification)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Theoretical stance: categorical vs dimensional ratings",
       x = "Categorical rating (0–100)",
       y = "Dimensional rating (0–100)",
       colour = "Classification") +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100))
```

## Operational stance

```{r operational-scatter}
df |>
  ggplot(aes(x = operational_stance_categorical_rating,
             y = operational_stance_dimensional_rating,
             colour = overall_classification_classification)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Operational stance: categorical vs dimensional ratings",
       x = "Categorical rating (0–100)",
       y = "Dimensional rating (0–100)",
       colour = "Classification") +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100))
```

## Theory vs operations: do they diverge?

```{r theory-vs-ops}
divergence <- df |>
  mutate(
    theory_net = theoretical_stance_dimensional_rating - theoretical_stance_categorical_rating,
    ops_net    = operational_stance_dimensional_rating - operational_stance_categorical_rating
  )

divergence |>
  ggplot(aes(x = ops_net, y = theory_net,
             colour = overall_classification_classification)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Theory–operations divergence",
       subtitle = "Positive = more dimensional, Negative = more categorical",
       x = "Operational net (dim − cat)",
       y = "Theoretical net (dim − cat)",
       colour = "Classification") +
  coord_equal()
```


# 5. Ratings by Domain (Sampling, Data Handling, Analysis)

```{r domain-ratings-long}
domains <- df |>
  select(filename, overall_classification_classification,
         starts_with("methods_extraction_sampling_"),
         starts_with("methods_extraction_data_handling_"),
         starts_with("methods_extraction_primary_analysis_")) |>
  pivot_longer(
    cols = -c(filename, overall_classification_classification),
    names_to = "field",
    values_to = "value"
  ) |>
  mutate(
    domain = case_when(
      str_detect(field, "sampling")         ~ "Sampling",
      str_detect(field, "data_handling")     ~ "Data Handling",
      str_detect(field, "primary_analysis")  ~ "Analysis Technique"
    ),
    measure = case_when(
      str_detect(field, "categorical_rating") ~ "Categorical",
      str_detect(field, "dimensional_rating") ~ "Dimensional",
      str_detect(field, "confidence")         ~ "Confidence",
      TRUE                                    ~ "technique"
    )
  ) |>
  filter(measure %in% c("Categorical", "Dimensional"))
```

```{r domain-boxplots}
domains |>
  ggplot(aes(x = domain, y = value, fill = measure)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_manual(values = c("Categorical" = "#e41a1c", "Dimensional" = "#377eb8")) +
  labs(title = "Categorical vs dimensional ratings by methodological domain",
       x = NULL, y = "Rating (0–100)", fill = NULL) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r domain-faceted}
domains |>
  ggplot(aes(x = value, fill = measure)) +
  geom_histogram(bins = 20, alpha = 0.6, position = "identity") +
  facet_wrap(~domain, scales = "free_y") +
  scale_fill_manual(values = c("Categorical" = "#e41a1c", "Dimensional" = "#377eb8")) +
  labs(title = "Distribution of ratings by domain",
       x = "Rating (0–100)", y = "Count", fill = NULL)
```


# 6. Ratings Over Time

```{r ratings-over-time}
time_df <- df |>
  filter(!is.na(year)) |>
  group_by(year) |>
  summarise(
    n = n(),
    mean_dim_theory = mean(theoretical_stance_dimensional_rating, na.rm = TRUE),
    mean_cat_theory = mean(theoretical_stance_categorical_rating, na.rm = TRUE),
    mean_dim_ops    = mean(operational_stance_dimensional_rating, na.rm = TRUE),
    mean_cat_ops    = mean(operational_stance_categorical_rating, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n >= 3)  # only plot years with enough data

time_long <- time_df |>
  pivot_longer(starts_with("mean_"), names_to = "measure", values_to = "rating") |>
  mutate(
    stance   = if_else(str_detect(measure, "theory"), "Theoretical", "Operational"),
    direction = if_else(str_detect(measure, "_dim_"), "Dimensional", "Categorical")
  )

time_long |>
  ggplot(aes(x = year, y = rating, colour = direction, linetype = stance)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Categorical" = "#e41a1c", "Dimensional" = "#377eb8")) +
  labs(title = "Mean dimensional vs categorical ratings over time",
       x = "Year", y = "Mean rating (0–100)",
       colour = NULL, linetype = NULL)
```

```{r classification-over-time}
df |>
  filter(!is.na(year)) |>
  count(year, overall_classification_classification) |>
  group_by(year) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = year, y = prop, fill = overall_classification_classification)) +
  geom_area(alpha = 0.8) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent) +
  labs(title = "Classification proportions over time",
       x = "Year", y = "Proportion", fill = "Classification")
```


# 7. Methods Techniques

```{r sampling-technique}
df |>
  count(methods_extraction_sampling_technique, sort = TRUE) |>
  filter(methods_extraction_sampling_technique != "N/A") |>
  ggplot(aes(x = reorder(methods_extraction_sampling_technique, n), y = n)) +
  geom_col(fill = "#756bb1") +
  coord_flip() +
  labs(title = "Sampling techniques", x = NULL, y = "Count")
```

```{r analysis-technique}
df |>
  count(methods_extraction_primary_analysis_technique, sort = TRUE) |>
  filter(methods_extraction_primary_analysis_technique != "N/A") |>
  ggplot(aes(x = reorder(methods_extraction_primary_analysis_technique, n), y = n)) +
  geom_col(fill = "#de2d26") +
  coord_flip() +
  labs(title = "Primary analysis techniques", x = NULL, y = "Count")
```

```{r data-handling}
df |>
  count(methods_extraction_data_handling_technique, sort = TRUE) |>
  filter(methods_extraction_data_handling_technique != "N/A") |>
  ggplot(aes(x = reorder(methods_extraction_data_handling_technique, n), y = n)) +
  geom_col(fill = "#fc9272") +
  coord_flip() +
  labs(title = "Data handling techniques", x = NULL, y = "Count")
```


# 8. Construct-Level Analysis

```{r construct-direction}
constructs |>
  count(operational_direction, theoretical_direction) |>
  ggplot(aes(x = operational_direction, y = theoretical_direction, size = n, colour = n)) +
  geom_point() +
  scale_size_continuous(range = c(3, 20)) +
  scale_colour_viridis_c() +
  labs(title = "Construct-level: operational vs theoretical direction",
       x = "Operational direction", y = "Theoretical direction",
       size = "Count", colour = "Count")
```

```{r personality-disorders}
pd_counts <- constructs |>
  filter(personality_disorder == TRUE) |>
  count(construct_label_normalised, sort = TRUE)

if (nrow(pd_counts) > 0) {
  pd_counts |>
    ggplot(aes(x = reorder(construct_label_normalised, n), y = n)) +
    geom_col(fill = "#e6550d") +
    coord_flip() +
    labs(title = "Personality disorder constructs", x = NULL, y = "Count")
} else {
  cat("No personality disorder constructs found in this sample.\n")
}
```


# 9. Evidence & Discourse

```{r evidence-direction}
evidence <- detail_df |>
  filter(str_detect(detail_table, "^evidence_extraction"))

evidence |>
  count(detail_table, sort = TRUE) |>
  mutate(direction = str_remove(detail_table, "evidence_extraction\\.")) |>
  ggplot(aes(x = reorder(direction, n), y = n)) +
  geom_col(fill = "#31a354") +
  coord_flip() +
  labs(title = "Evidence pieces by direction",
       x = NULL, y = "Count")
```

```{r evidence-location}
evidence |>
  filter(!is.na(location)) |>
  count(location, sort = TRUE) |>
  ggplot(aes(x = reorder(location, n), y = n)) +
  geom_col(fill = "#3182bd") +
  coord_flip() +
  labs(title = "Evidence by article section", x = NULL, y = "Count")
```

```{r discourse-types}
discourse <- detail_df |>
  filter(detail_table == "discourse.key_passages")

if (nrow(discourse) > 0) {
  discourse |>
    count(type, sort = TRUE) |>
    ggplot(aes(x = reorder(type, n), y = n)) +
    geom_col(fill = "#636363") +
    coord_flip() +
    labs(title = "Discourse types", x = NULL, y = "Count")
}
```


# 10. Frameworks Cited

```{r frameworks}
# Parse semicolon-separated framework citation fields
parse_frameworks <- function(df, col_name, label) {
  df |>
    select(filename, value = all_of(col_name)) |>
    filter(!is.na(value), value != "N/A", value != "") |>
    separate_rows(value, sep = ";\\s*") |>
    mutate(value = str_trim(value)) |>
    filter(value != "", value != "N/A") |>
    count(value, sort = TRUE) |>
    mutate(citation_type = label)
}

framework_cols <- c(
  "theoretical_extraction_positive_framework_citations",
  "theoretical_extraction_negative_framework_citations",
  "theoretical_extraction_neutral_framework_citations"
)

# Only plot if columns exist
if (all(framework_cols %in% names(df))) {
  frameworks <- bind_rows(
    parse_frameworks(df, framework_cols[1], "Positive"),
    parse_frameworks(df, framework_cols[2], "Negative"),
    parse_frameworks(df, framework_cols[3], "Neutral")
  )

  if (nrow(frameworks) > 0) {
    frameworks |>
      group_by(citation_type) |>
      slice_max(n, n = 10) |>
      ungroup() |>
      ggplot(aes(x = reorder(value, n), y = n, fill = citation_type)) +
      geom_col() +
      coord_flip() +
      facet_wrap(~citation_type, scales = "free_y") +
      scale_fill_manual(values = c("Positive" = "#2ca02c", "Negative" = "#d62728", "Neutral" = "#7f7f7f")) +
      labs(title = "Top framework citations by valence",
           x = NULL, y = "Count") +
      theme(legend.position = "none")
  }
}
```


# 11. Summary Table

```{r summary-table}
summary_stats <- df |>
  summarise(
    `Total articles`       = n(),
    `Categorical`          = sum(overall_classification_classification == "categorical", na.rm = TRUE),
    `Dimensional`          = sum(overall_classification_classification == "dimensional", na.rm = TRUE),
    `Hybrid`               = sum(overall_classification_classification == "hybrid", na.rm = TRUE),
    `Indeterminate`        = sum(overall_classification_classification == "indeterminate", na.rm = TRUE),
    `Mean dim (theory)`    = round(mean(theoretical_stance_dimensional_rating, na.rm = TRUE), 1),
    `Mean cat (theory)`    = round(mean(theoretical_stance_categorical_rating, na.rm = TRUE), 1),
    `Mean dim (ops)`       = round(mean(operational_stance_dimensional_rating, na.rm = TRUE), 1),
    `Mean cat (ops)`       = round(mean(operational_stance_categorical_rating, na.rm = TRUE), 1),
    `Mean confidence`      = round(mean(overall_classification_confidence, na.rm = TRUE), 2)
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value")

knitr::kable(summary_stats, caption = "Summary statistics")
```
